---
title: "Wine Quality EDA"
Names: Safina Davis, Mita Crane, Mackenzie Steinbiss
output: html_document
date: "2025-05-01"
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Business Understanding

##What Makes a High-Quality Wine?

Grape Variety and Terroir Grape Variety: Different grape varieties
possess unique characteristics in terms of flavor, aroma, acidity, and
tannin levels. These traits, combined with the winemaker’s skill,
contribute to the overall style and quality of the wine.

Terroir: This French term encompasses the environmental factors that
affect a crop's phenotype, including unique environment contexts,
farming practices, and a crop's specific growth habitat. Terroir
includes climate, soil, topography, and other factors that influence the
grape's growth and, consequently, the wine's quality.

2.  Viticultural Practices Canopy Management: Proper leaf and shoot
    management ensures optimal sunlight exposure and air circulation,
    reducing disease risk and promoting even ripening.

Yield Control: Limiting the number of grape clusters can lead to more
concentrated flavors, enhancing wine quality.

Harvest Timing: Picking grapes at the right time ensures a balance
between sugar, acidity, and phenolic maturity.

3.  Winemaking Techniques Fermentation Control: Managing fermentation
    temperature and duration affects the development of desired flavors
    and aromas.

Use of Oak: Aging wine in oak barrels can impart additional flavors and
tannins, contributing to complexity.

Malolactic Fermentation: This secondary fermentation can soften acidity
and add buttery notes to the wine.

4.  Chemical Composition Alcohol Content: Higher alcohol levels can
    enhance body and mouthfeel but must be balanced to avoid
    overpowering other flavors.

Acidity: Acidity provides freshness and structure; too little can make
wine taste flat, while too much can make it sharp.

Tannins: These compounds add structure and astringency, important for
red wines and aging potential.

Residual Sugar: The amount of sugar left after fermentation affects
sweetness and balance.

5.  Sensory Attributes Aroma and Flavor Complexity: High-quality wines
    exhibit a range of aromas and flavors that evolve over time.

Balance: A harmonious integration of all components—acidity, tannins,
alcohol, and sweetness—is crucial.

Finish: The length of time flavors linger after swallowing; a longer
finish is often associated with higher quality.

6.  Aging Potential Wines with the right balance of acidity, tannins,
    and fruit concentration can develop more complex flavors over time,
    enhancing quality.

7.  Absence of Faults High-quality wines are free from defects such as
    oxidation, cork taint, or excessive sulfur dioxide, which can
    negatively impact flavor and aroma.

##Here I am loading the dataset for white wine into R and seperating the
commas into columns, and loading necessary libraries

```{r}
setwd("C:/Users/19177/OneDrive/Desktop")


White_wine <- read.csv("winequality-white.csv", sep = ";")
head(White_wine)
# Install required packages if not already installed
if(!require(tidyverse)) install.packages("tidyverse")
if(!require(corrplot)) install.packages("corrplot")
if(!require(ggplot2)) install.packages("ggplot2")
if(!require(dplyr)) install.packages("dplyr")
if(!require(knitr)) install.packages("knitr")
if(!require(gridExtra)) install.packages("gridExtra")
if(!require(car)) install.packages("car")
if(!require(reshape2)) install.packages("reshape2")

# Load the libraries
library(tidyverse)
library(corrplot)
library(ggplot2)
library(dplyr)
library(knitr)
library(gridExtra)
library(car)
library(reshape2)
```

##Loading and combining datasets

```{r}
# Set your working directory if needed
setwd("C:/Users/19177/OneDrive/Desktop")
# Load the datasets
white_wine <- read.csv("winequality-white.csv", sep = ";")
red_wine <- read.csv("winequality-red.csv", sep = ";")

# Add a type variable to distinguish between red and white wine
white_wine$type <- "white"
red_wine$type <- "red"

# Combine both datasets
wine_data <- rbind(white_wine, red_wine)

# Examine the structure of the combined dataset
str(wine_data)
```

# Data Understanding

When was the data acquired? - The UCI Wine Quality dataset does not
specify exact collection dates. However, it was made publicly available
through the UCI Machine Learning Repository and has been widely used
since at least 2009. It is reasonable to assume that the data was
collected sometime prior to that year.

Where was the data acquired? - The data was acquired from
physicochemical tests conducted by the Portuguese "Vinho Verde" Wine
Region. All wines originated from northern Portugal, specifically from
the Minho region.

How was the data acquired? - The dataset was created using
physicochemical (e.g., pH, alcohol) and sensory (quality scores) tests
on red and white variants of the Portuguese "Vinho Verde" wine. The
sensory quality scores were provided by at least three wine experts, and
the physicochemical properties were measured using standard analytical
chemistry techniques.

What are the attributes of this dataset? Description of each column in
the dataset:

fixed acidity - Concentration of non-volatile acids (mainly tartaric).
Helps preserve wine. volatile acidity - Amount of acetic acid. High
levels give wine a vinegar taste. citric acid - A weak organic acid that
adds freshness and flavor. residual sugar - Sugar left after
fermentation; wines with \>45g/L are considered sweet. chlorides - Salt
content in wine. High values can indicate undesirable tastes. free
sulfur dioxide - SO₂ in free form; acts as antioxidant and antimicrobial
agent. total sulfur dioxide - Total SO₂, including both free and bound
forms. Excess may affect taste. density - Mass-to-volume ratio; can
relate to sugar and alcohol content. pH - Indicates acidity/basicity.
Affects stability, color, and taste. sulphates - Additive for
stabilization and as an antimicrobial. Higher levels may be perceived as
bitter. alcohol - Alcohol percentage by volume. Often associated with
wine body and warmth. quality - Quality score (0–10) based on sensory
analysis by experts. This is the target variable.

What type of data do these attributes contain?

fixed acidity Numerical (continuous, ratio) volatile acidity Numerical
(continuous, ratio) citric acid Numerical (continuous, ratio) residual
sugar Numerical (continuous, ratio) chlorides Numerical (continuous,
ratio) free sulfur dioxide Numerical (continuous, ratio) total sulfur
dioxide Numerical (continuous, ratio) density Numerical (continuous,
ratio) pH Numerical (continuous, interval) sulphates Numerical
(continuous, ratio) alcohol Numerical (continuous, ratio) quality
Ordinal (integers from 0 to 10)

#Data exploration Summary statistics and check for missing values

```{r}
# Summary statistics
summary(wine_data)

# Check for missing values
colSums(is.na(wine_data))

# Get dimensions of datasets
cat("White wine dataset dimensions:", dim(white_wine), "\n")
cat("Red wine dataset dimensions:", dim(red_wine), "\n")
cat("Combined dataset dimensions:", dim(wine_data), "\n")

# Check class distribution (quality) in both datasets
table(white_wine$quality)
table(red_wine$quality)
table(wine_data$quality)
```

Converting wine quality into a factor for analysis

```{r}
# Convert quality to a factor for classification purposes
wine_data$quality_category <- ifelse(wine_data$quality >= 7, "high", 
                               ifelse(wine_data$quality <= 4, "low", "medium"))
wine_data$quality_category <- factor(wine_data$quality_category, 
                                    levels = c("low", "medium", "high"))

# Check the distribution of quality categories
table(wine_data$quality_category)
table(wine_data$type, wine_data$quality_category)
```

Examining distribution of values with histograms

```{r}
# Function to create histogram for a specific variable
create_histogram <- function(data, variable, binwidth = NULL, title = NULL) {
  if(is.null(title)) title <- paste("Distribution of", variable)
  
  ggplot(data, aes_string(x = variable, fill = "type")) +
    geom_histogram(position = "dodge", alpha = 0.7, binwidth = binwidth) +
    labs(title = title, x = variable, y = "Count") +
    theme_minimal() +
    scale_fill_manual(values = c("red" = "#E74C3C", "white" = "#F7DC6F"))
}

# Create histograms for key variables
p1 <- create_histogram(wine_data, "alcohol", binwidth = 0.2)
p2 <- create_histogram(wine_data, "pH", binwidth = 0.05)
p3 <- create_histogram(wine_data, "fixed.acidity", binwidth = 0.5)
p4 <- create_histogram(wine_data, "volatile.acidity", binwidth = 0.05)

# Display plots in a grid
grid.arrange(p1, p2, p3, p4, ncol = 2)
```

Quality distribution by wine type

```{r}
# Visualize the distribution of quality scores by wine type
ggplot(wine_data, aes(x = factor(quality), fill = type)) +
  geom_bar(position = "dodge", alpha = 0.7) +
  labs(title = "Distribution of Quality Scores by Wine Type",
       x = "Quality Score", y = "Count") +
  theme_minimal() +
  scale_fill_manual(values = c("red" = "#E74C3C", "white" = "#F7DC6F"))
```

Correlation analysis

```{r}
# Function to create correlation matrix
plot_correlation <- function(data, title) {
  # Remove non-numeric columns
  data_numeric <- data %>% select_if(is.numeric)
  
  # Calculate correlation matrix
  corr_matrix <- cor(data_numeric)
  
  # Plot correlation matrix
  corrplot(corr_matrix, method = "color", type = "upper", order = "hclust",
           tl.col = "black", tl.srt = 45, addCoef.col = "black", 
           number.cex = 0.7, title = title, mar = c(0, 0, 1, 0))
}

# Plot correlation matrices
par(mfrow = c(1, 1))
plot_correlation(wine_data, "Correlation Matrix - All Wines")
```

Analyzing quality according to key features

```{r}
# Function to create boxplots
create_boxplot <- function(data, y_var, title = NULL) {
  if(is.null(title)) title <- paste(y_var, "vs. Quality by Wine Type")
  
  ggplot(data, aes_string(x = "factor(quality)", y = y_var, fill = "type")) +
    geom_boxplot(alpha = 0.7) +
    labs(title = title, x = "Quality Score", y = y_var) +
    theme_minimal() +
    scale_fill_manual(values = c("red" = "#E74C3C", "white" = "#F7DC6F"))
}

# Create boxplots for key variables vs quality
p5 <- create_boxplot(wine_data, "alcohol")
p6 <- create_boxplot(wine_data, "volatile.acidity")
p7 <- create_boxplot(wine_data, "sulphates")
p8 <- create_boxplot(wine_data, "total.sulfur.dioxide")

# Display boxplots in a grid
grid.arrange(p5, p6, p7, p8, ncol = 2)
```

Statistical analysis of red and white wine differences

```{r}
# T-tests for key variables between red and white wines
t_test_results <- data.frame(
  Variable = character(),
  t_value = numeric(),
  p_value = numeric(),
  significant = character(),
  stringsAsFactors = FALSE
)

variables_to_test <- c("fixed.acidity", "volatile.acidity", "citric.acid", "residual.sugar",
                      "chlorides", "free.sulfur.dioxide", "total.sulfur.dioxide", 
                      "density", "pH", "sulphates", "alcohol", "quality")

for(var in variables_to_test) {
  t_result <- t.test(red_wine[[var]], white_wine[[var]])
  t_test_results <- rbind(t_test_results, data.frame(
    Variable = var,
    t_value = round(t_result$statistic, 3),
    p_value = round(t_result$p.value, 5),
    significant = ifelse(t_result$p.value < 0.05, "Yes", "No"),
    stringsAsFactors = FALSE
  ))
}

# Display t-test results
kable(t_test_results, caption = "T-test Results: Red vs. White Wine")
```

Linear regression analysis

```{r}
# setwd("path/to/your/data")
setwd("C:/Users/19177/OneDrive/Desktop")
# Load the datasets
white_wine <- read.csv("winequality-white.csv", sep = ";")
red_wine <- read.csv("winequality-red.csv", sep = ";")

# Add type labels
red_wine$type <- "red"
white_wine$type <- "white"

# Combine into one dataset
wine_data <- rbind(red_wine, white_wine)
# Fit linear regression model for quality
model_all <- lm(quality ~ . - type, data = wine_data)

model_summary <- summary(model_all)

# Display regression results
model_summary

# Check for multicollinearity
vif_results <- vif(model_all)
vif_data <- data.frame(
  Variable = names(vif_results),
  VIF = round(vif_results, 2),
  stringsAsFactors = FALSE
)
kable(vif_data, caption = "Variance Inflation Factors")

# Fit separate models for red and white wines


model_red <- lm(quality ~ . , data = red_wine[, !(names(red_wine) %in% c("type"))])

model_white <- lm(quality ~ . , data = white_wine[, !(names(white_wine) %in% c("type"))])


# Compare key coefficients
coef_comparison <- data.frame(
  Variable = names(coef(model_all))[2:length(coef(model_all))],
  All_Wines = round(coef(model_all)[2:length(coef(model_all))], 4),
  Red_Wine = round(coef(model_red)[2:length(coef(model_red))], 4),
  White_Wine = round(coef(model_white)[2:length(coef(model_white))], 4),
  stringsAsFactors = FALSE
)

kable(coef_comparison, caption = "Comparison of Regression Coefficients")
```

Key differences by wine type

```{r}
# Calculate mean values for key variables by wine type
wine_means <- wine_data %>%
  group_by(type) %>%
  summarize(across(c(fixed.acidity, volatile.acidity, citric.acid, residual.sugar,
                    chlorides, free.sulfur.dioxide, total.sulfur.dioxide, 
                    density, pH, sulphates, alcohol, quality), mean))

# Reshape data for visualization
wine_means_long <- melt(wine_means, id.vars = "type")

# Plot mean values by wine type
ggplot(wine_means_long, aes(x = variable, y = value, fill = type)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Mean Values of Chemical Properties by Wine Type", 
       x = "Variable", y = "Mean Value") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_fill_manual(values = c("red" = "#E74C3C", "white" = "#F7DC6F"))
```
